---
title: "PMMH Application to Stochastic Volatility Model"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, cache = TRUE)
library(Rcpp)
library(tidyverse)
library(latex2exp)
library(SMC)
```

## Simulate Dataset

```{r include = F}
tmax <- 500
param <- c(0.9, 1, 1) # True model paramters.
```

Generate dataset
```{r eval = F}
set.seed(192837465)
data <- SMC::stochastic_volatility(tmax, param)
# saveRDS(data,"Data/sv_example_data.RDS") # save dataset.
```


## PMMH settings used for producing the plots.
* niter = 800
* N = 300
* alpha_init: 0.3
* alpha_prop: beta(1, 100, 100/thetagiven - 100)
* alpha_prior: beta(6, 6/0.8 - 6)
* sigma_init: 0.5
* sigma_prop: gamma(1, 200, thetagiven/200)
* sigma_prior: gamma(theta, 5, 0.5/5

Run the algorithm with the settings described above and save the output for future use.

```{r eval = F}
N <- 300
initial <- c(0.3, 1, 0.5)
mh_results <- SMC::pmmh1(thetastart = initial[3],
                   niter=800,
                   N=N,
                   y=y,
                   burnin=0,
                   alpha = initial[1],
                   beta = param[2])

#saveRDS(mh_results, 'Data/mh_results.RDS')
```

Load the saveddataset and MH output.
```{r}
data <- readRDS("Data/sv_example_data.RDS")
y <- data[-1,2]
mh_results <- readRDS("Data/mh_results.RDS")
```


```{r echo=FALSE}
df <- data.frame(states=1:dim(data)[1],x=data[,1],y=data[,2])
ggplot(data = df)+
  geom_line(aes(x=states,y=x),color = "black", size = 0.4)+
  geom_point(aes(x=states, y=y), color='grey40', alpha=0.5,size=0.3)+
  labs(x="time",y = "")+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 
# ggsave("Figures/true_x_y.png",height = 5, width=8,units = "cm")
```

## State estimation using the real parameters

APF
```{r}
bf_result <- SMC::BSF(y,N, param)
#apf_state_estimates <- apf_result$states[-1]
data.frame(state = 1:length(bf_result$states[-1])[1],
                 state_estimate=bf_result$states[-1],
                 real_state = data[-1,1]) %>% 
  ggplot()+
    geom_line(aes(x=state,y=state_estimate), color = '#1ea896', size=0.4)+
    geom_line(aes(x=state,y=real_state), size=0.4)+
    labs(x="time",y="state")+
    theme(panel.grid = element_blank(),
          panel.background = element_blank(),
          text = element_text(family="LM Roman 10", size=10),
          axis.line=element_line(size = 0.3),
          axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
          axis.text = element_text(colour = "black"),
          aspect.ratio = 0.5,
          plot.margin=grid::unit(c(0,0.3,0,0), "cm")
          ) 
```

APF
```{r}
bf_result <- SMC::BSF(y,N, param)
apf_result <- SMC::APF(y, N, param)
#apf_state_estimates <- apf_result$states[-1]
data.frame(state = 1:length(apf_result$states[-1])[1],
                state_estimate_bf=bf_result$states[-1],
                 state_estimate_apf=apf_result$states[-1],
                 real_state = data[-1,1]) %>% 
  ggplot()+
    geom_line(aes(x=state,y=state_estimate_bf), color = '#1ea896', size=0.4)+
    geom_line(aes(x=state,y=state_estimate_apf), color = '#ee6c4d', size=0.4)+
    geom_line(aes(x=state,y=real_state), size=0.4)+
    labs(x="time",y="state")+
    theme(panel.grid = element_blank(),
          panel.background = element_blank(),
          text = element_text(family="LM Roman 10", size=10),
          axis.line=element_line(size = 0.3),
          axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
          axis.text = element_text(colour = "black"),
          aspect.ratio = 0.5,
          plot.margin=grid::unit(c(0,0.3,0,0), "cm")
          ) 
```


Compute the mean of the parameter estimates from iteration $200$ to $800$.
```{r}
# sigma
mean(mh_results$samples[200:800])

# alpha
mean(mh_results$alphasamples[200:800])
```

### Plots of the states and observations
```{r , echo=FALSE}
df <- data.frame(states=1:dim(data)[1],x=data[,1],y=data[,2])
ggplot(data = df)+
  geom_line(aes(x=states,y=x),color = "black", size = 0.4)+
  labs(x="time",y = "x")+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 
# ggsave("Figures/true_x.png",height = 5, width=8,units = "cm")
```

```{r echo=FALSE}
ggplot(data = df)+
  geom_line(aes(x=states,y=y),color = "black", size = 0.4)+
  labs(x="time",y = "y")+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 
# ggsave("Figures/true_y.png",height = 5, width=8,units = "cm")
```


### Trace plots
```{r, echo=FALSE}
samples <- data.frame(iteration = 1:length(mh_results$samples),
                      alpha=mh_results$alphasamples,
                      sigma=mh_results$samples)

ggplot(data = samples, aes(x = iteration, y = alpha))+
  geom_line(color = "#264653", size = 0.4)+
  geom_hline(yintercept = param[1],linetype = 'dashed', size = 0.4)+
  labs(y = expression(alpha))+
  scale_y_continuous(limits=c(0,1),breaks=c(seq(0,1,by=0.2),param[1]),labels=c(seq(0,1,by=0.2),param[1]))+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 

# ggsave("Figures/trace_alpha.png",height = 5, width=8,units = "cm")

ggplot(data = samples, aes(x = iteration, y = sigma))+
  geom_line(color = "#e76f51", size = 0.4)+
  geom_hline(yintercept = param[3], linetype = 'dashed', size = 0.4)+
  labs(y = expression(sigma))+
  scale_y_continuous(limits=c(0.5,2.5),breaks=seq(0.5,2.5,by=0.25),labels=seq(0.5,2.5,by=0.25))+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 

# ggsave("Figures/trace_sigma.png",height = 5, width=8,units = "cm")
```

### Posterior estimates
```{r, echo=FALSE}
ggplot(data = samples)+
  geom_histogram(aes(alpha), bins = 25, color = "#264653", fill="#264653",size = 0.4,alpha=0.4)+
  geom_vline(xintercept = param[1],linetype = 'dashed',size = 0.4)+
  labs(x = expression(alpha), y="")+
  scale_x_continuous(limits=c(0,1),breaks=c(seq(0,1,by=0.2),param[1]),labels=c(seq(0,1,by=0.2),param[1]))+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 
# ggsave("Figures/posterior_alpha.png",height = 5, width=8,units = "cm")

ggplot(data = samples)+
  geom_histogram(aes(sigma), bins=25, color = "#e76f51", fill="#e76f51",size = 0.4,alpha=0.4)+
  geom_vline(xintercept = param[3],linetype = 'dashed',size = 0.4)+
  labs(x = expression(sigma),y="")+
  scale_x_continuous(limits=c(0.5,2.5),breaks=c(seq(0.5,2.5,by=0.25),param[3]),labels=c(seq(0.5,2.5,by=0.25),param[3]))+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 
# ggsave("Figures/posterior_sigma.png",height = 5, width=8,units = "cm")
```

### Plots of state estimates at different stages of the PMMH

State estimates based on the the initial parameter values.
```{r, echo=FALSE}
# x_filtered_init <- SMC::BSF(y, N, initial)$states[-1,]
#saveRDS(x_filtered_init, "Data/x_filtered_init.RDS")
x_filtered_init <- readRDS("Data/x_filtered_init.RDS")

data.frame(state = 1:dim(mh_results$final_states)[1],
                 state_estimate=x_filtered_init,
                 real_state = data[,1]) %>% 
  ggplot()+
    geom_line(aes(x=state,y=state_estimate), color = '#1ea896', size=0.4)+
    geom_line(aes(x=state,y=real_state), size=0.4)+
    labs(x="time",y="state")+
    theme(panel.grid = element_blank(),
          panel.background = element_blank(),
          text = element_text(family="LM Roman 10", size=10),
          axis.line=element_line(size = 0.3),
          axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
          axis.text = element_text(colour = "black"),
          aspect.ratio = 0.5,
          plot.margin=grid::unit(c(0,0.3,0,0), "cm")
          ) 

# ggsave("Figures/states_init.png",height = 5, width=8,units = "cm")
```

State estimates based on the particle estimates at iteration $50$.
```{r, echo=FALSE}
# param_iter_50 <- c(samples$alpha[50], 1, samples$sigma[50])
# x_filtered_at_iter_50 <- SMC::BSF(y, N, param_iter_50)$states[-1,]
# saveRDS(x_filtered_at_iter_50, "Data/x_filtered_at_iter_50.RDS")

x_filtered_at_iter_50 <- readRDS("Data/x_filtered_at_iter_50.RDS")

data.frame(state = 1:dim(mh_results$final_states)[1],
                 state_estimate=x_filtered_at_iter_50,
                 real_state = data[,1]) %>% 
  ggplot()+
  geom_line(aes(x=state,y=state_estimate), color = '#1ea896')+
  geom_line(aes(x=state,y=real_state))+
  labs(x="time",y="state")+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 

# ggsave("Figures/states_50.png",height = 5, width=8,units = "cm")
```

State estimates based on the particle estimates at iteration $100$.
```{r, echo=FALSE}
# param_iter_100 <- c(samples$alpha[100], 1, samples$sigma[100])
# x_filtered_at_iter_100 <- SMC::BSF(y, N, param_iter_100)$states[-1,]
# saveRDS(x_filtered_at_iter_100, "Data/x_filtered_at_iter_100.RDS")

x_filtered_at_iter_100 <- readRDS("Data/x_filtered_at_iter_100.RDS")


data.frame(state = 1:dim(mh_results$final_states)[1],
                 state_estimate=x_filtered_at_iter_100,
                 real_state = data[,1]) %>% 
  ggplot()+
  geom_line(aes(x=state,y=state_estimate), color = '#1ea896')+
  geom_line(aes(x=state,y=real_state))+
    labs(x="time",y="state")+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 
# ggsave("Figures/states_100.png",height = 5, width=8,units = "cm")
```

State estimates based on the final parameter estimates.
```{r, echo=FALSE}
data.frame(state = 1:dim(mh_results$final_states)[1],
                 mean_state_estimate=rowMeans(mh_results$final_states),
                 real_state = data[,1]) %>% 
  ggplot()+
  geom_line(aes(x=state,y=mean_state_estimate), color = '#1ea896')+
  geom_line(aes(x=state,y=real_state))+
    labs(x="time",y="state")+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        )  
# ggsave("Figures/states_final.png",height = 5, width=8,units = "cm")
```


```{r, echo=FALSE}
ggplot.corr <- function(data, lag.max = 100, ci = 0.95, large.sample.size = TRUE, horizontal = TRUE,...) {
  
  if(horizontal == TRUE) {numofrow <- 1} else {numofrow <- 2}
  
  list.acf <- acf(data, lag.max = lag.max, type = "correlation", plot = FALSE)
  N <- as.numeric(list.acf$n.used)
  df1 <- data.frame(lag = list.acf$lag, acf = list.acf$acf)
  df1$lag.acf <- dplyr::lag(df1$acf, default = 0)
  df1$lag.acf[2] <- 0
  df1$lag.acf.cumsum <- cumsum((df1$lag.acf)^2)
  df1$acfstd <- sqrt(1/N * (1 + 2 * df1$lag.acf.cumsum))
  df1$acfstd[1] <- 0
  df1 <- select(df1, lag, acf, acfstd)
  
  list.pacf <- acf(data, lag.max = lag.max, type = "partial", plot = FALSE)
  df2 <- data.frame(lag = list.pacf$lag,pacf = list.pacf$acf)
  df2$pacfstd <- sqrt(1/N)
    plot.acf <- ggplot(data = df1, aes( x = lag, y = acf)) +
    geom_col(color="#264653", width = 0.01) +
    geom_hline(yintercept = qnorm((1+ci)/2)/sqrt(N), 
               colour = "black",
               linetype = "dashed",
               size=0.4) + 
    geom_hline(yintercept = - qnorm((1+ci)/2)/sqrt(N), 
               colour = "black",
               linetype = "dashed",
               size=0.4) + 
    scale_x_continuous(breaks = seq(0,max(df1$lag),10)) +
    scale_y_continuous(name = element_blank(), 
                       limits = c(min(df1$acf,df2$pacf),1)) +
    theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(family="LM Roman 10", size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        )
  plot.acf
}
```


## Auto-correlation plots
```{r , echo=FALSE}
## compute acfs
acf_alpha <- acf(samples$alpha, plot = FALSE, lag.max=100)
acf_sigma <- acf(samples$sigma, plot = FALSE, lag.max=100)
```

ACF for $\alpha$

```{r echo=FALSE}
ggplot.corr(samples$alpha,large.sample.size = FALSE)
# ggsave("Figures/acf_alpha.png",height = 5, width=8,units = "cm")
```

ACF for $\sigma$

```{r echo=FALSE}
ggplot.corr(samples$sigma,large.sample.size = FALSE)
# ggsave("Figures/acf_sigma.png",height = 5, width=8,units = "cm")
```

