---
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, cache = TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = F)
```

```{r include = F}
library(SMC)
library(tidyverse)
library(latex2exp)
library(gridExtra)
```

As a part of our project we have written an R package, `SMC`, which is mostly
written in C++ using the `Rcpp/RcppArmadillo` package. In this tutorial we
provide an overview of the different algorithms implemented in the `SMC` 
package. For a description of the algorithms used within this tutorial, we refer
the reader to the primary report.

The algorithms within `SMC` have been written in order to conduct analysis on
data which can be modelled by a Stochastic Volatility Model (SVM):
\begin{equation}
X_t = \alpha X_{t-1} + \epsilon_t, \quad \epsilon \sim \mathcal{N}(0,1)
\end{equation}
\begin{equation}
Y_t = \beta \text{exp}(X_t / 2) \zeta_t, \quad \zeta_t \sim \mathcal{N}(0,1)
\end{equation}
We start by defining our model parameters $\theta = (\alpha, \beta, \sigma)$ and
generating synthetic data from the associated SVM.

```{r}
theta <- c(0.91, 1, 1)
tmax <- 1000
set.seed(1234)
data <- SMC::stochastic_volatility(tmax, theta)
head(data)
```

```{r include=F}
data_pl <- ggplot(data = cbind("t" = 1:tmax, data)) +
           geom_line(aes(x = t, y = x), col = "black", size = 0.4) +
           geom_point(aes(x = t, y = y), col = "grey40", alpha = 0.5, size = 0.3) +
           labs(x = "time", y = " ") +
           theme(panel.grid = element_blank(),
                 panel.background = element_blank(),
                 text = element_text(size=10),
                 axis.line=element_line(size = 0.3),
                 axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
                 axis.text = element_text(colour = "black"),
                 aspect.ratio = 0.5,
                 plot.margin=grid::unit(c(0,0.3,0,0), "cm"),
                )
pdf("Figures/Tutorial/synth_data.pdf")
data_pl
dev.off()
caption1 <- paste(
  "Synthetic data generated according to the Stochastic Volatility model with",
  "model parameters theta = (0.91, 1, 1)."
)
```

```{r, echo = F, fig.cap=caption1}
data_pl
```

Figure 1 shows a plot of the data generated using
`SMC::stochastic_volatility()`. One can clearly see that the volatility of the
observations, represented by the grey points, decreases as the states decrease.
In practice we do not have access to the unknown states $x_{0:t}$, however we 
can estimate them through the use of the **Bootstrap Partcile Filter** 
(BSF) and the **Auxiliary Particle Filter**.

```{r}
obs <- data$y
N <- 400 # We use 400 particles.
BSF_fit <- SMC::BSF(obs, N, theta) # Assumes we know the true model paramters.
APF_fit <- SMC::APF(obs, N, theta)

# Compute the MSE:
cat("Bootstrap MSE: ", sum((BSF_fit$states[-1] - data$x)^2), "\n",
    "Auxiliary MSE: ", sum((APF_fit$states[-1] - data$x)^2), sep = "")
```

```{r include = F}
filter_data <- data.frame(
  "t" = 1:tmax,
  "x" = data$x,
  "bsf" = BSF_fit$states[-1],
  "apf" = APF_fit$states[-1]
)
bsf_pl <- ggplot(data = filter_data) + 
          geom_line(aes(x = t, y = x), col = "black", size = 0.2) +
          geom_line(aes(x = t, y = bsf), col = "#1ea896", size = 0.2) +
          labs(x = "time", y = " ") +
          theme(panel.grid = element_blank(),
                panel.background = element_blank(),
                text = element_text(size=10),
                axis.line=element_line(size = 0.3),
                axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
                axis.text = element_text(colour = "black"),
                aspect.ratio = 0.5,
                plot.margin=grid::unit(c(0,0.3,0,0), "cm")
              )
apf_pl <- ggplot(data = filter_data) + 
          geom_line(aes(x = t, y = x), col = "black", size = 0.2) +
          geom_line(aes(x = t, y = apf), col = "#1ea896", size = 0.2) +
          labs(x = "time", y = " ") +
          theme(panel.grid = element_blank(),
                panel.background = element_blank(),
                text = element_text(size=10),
                axis.line=element_line(size = 0.3),
                axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
                axis.text = element_text(colour = "black"),
                aspect.ratio = 0.5,
                plot.margin=grid::unit(c(0,0.3,0,0), "cm")
              )
pdf("Figures/Tutorial/bsf_fit.pdf")
bsf_pl
pdf("Figures/Tutorial/apf_fit.pdf")
apf_pl
dev.off()
caption2 <- paste("Results of running running the BSF and APF (Left/Right) on the synthetic", 
                  "data. The true states are given by the solid black line,",
                  "and the filtered states are given by the dark green line.")
```

```{r, echo = F, fig.cap=caption2}
grid.arrange(bsf_pl, apf_pl, nrow = 1)
```

Figure 2 shows the filtered states plotted against the true synthetic states.
In this context there does not seem to be a huge difference in performance. We
have of course taken for granted that we know the true model parameters used to
generate the data. In practice we often must estimate the model parameters from 
the data. Working in the offline setting, we can do this using a Pseudo-Marginal
Metropolis Hastings (PMMH) algorithm. In short, one can use the above particle filters
to approximate the marinal likelihood of the observations
\begin{equation}
\hat{P}_N (y_{0:t} | \boldsymbol{\theta}) \approx p(y_{0:t} | \boldsymbol{\theta}).
\end{equation}
We can incorporate this approximation into the Metropolis Hastings algorithm,
and use it to accept/reject proposed values. For further details of this 
algorithm we once again refer the reader to the main report.

There are two versions of PMMH algorithm implemented within the `SMC` package.
The first, `pmmh1`, estimates parameters $\alpha$ and $\sigma$ **only**, using 
the prior/proposal
\begin{equation}
p(\alpha) = \text{Beta}(6, \frac{6}{0.8} - 6), \quad q(\alpha^*|\alpha) = 
\text{Beta}(100, \frac{100}{\alpha} - 100)
\end{equation}
\begin{equation}
p(\sigma) = \text{Gamma}(5, \frac{0.5}{5}), \quad q(\sigma^*|\sigma) = 
\text{Beta}(200, \frac{\sigma}{200}).
\end{equation}
The second, `pmmh2` estimates all three parameters $\theta = (\alpha, \beta, \sigma)$ 
and uses the prior
\begin{equation}
p(\alpha) = \text{Trunc-Normal}_{(-1,1)}(0.9, 0.5)
\end{equation}
\begin{equation}
p(\beta) = p(\sigma) = \text{Gamma}(2, 2)
\end{equation}
with the proposal
\begin{equation}
q(\theta^* | \theta) = \mathcal{N}(\theta, \sigma_{\text{prop}} \boldsymbol{I}).
\end{equation}
Since all of the parameters are bounded ($\sigma$ and $\beta$ must be positive; 
$\|\alpha\| < 1$), we have applied reflection at each of their respective
boundaries. This ensures proposals are only ever in pheasible regions. In this
tutorial we will show the use of `pmmh2`:

```{r eval = F}
pmmh2_fit <- SMC::pmmh2(10000, data$y, 800, c(0.5, 0.5, 0.5), 0.045)
```

This function outputs a list containing the number of accepted steps and a
matrix who's columns represent a markov chain for each parameter.

```{r include = F}
#saveRDS(pmmh2_fit, 'Data/Tutorial/pmmh2_fit.RDS')
pmmh2_fit <- readRDS("Data/Tutorial/pmmh2_fit.RDS")
```

```{r}
acceptance_rate <- pmmh2_fit$Accepted / nrow(pmmh2_fit$chain[-(1:2500),])
cat("Acceptance rate: ", acceptance_rate * 100, "%", sep = "")
```

The acceptance rate should ideally be around 23.4%. Let's examine the trace 
plots:

```{r include = F}
colours <- c("#264653","#e76f51","#B17D00")
tr_plot <- function(chains, i, colours) {
  df <- data.frame("t" = 1:nrow(chains), "y" = chains[,i])
  pl <- ggplot(data = df) +
        geom_line(aes(x = t, y = y), color = colours[i], size = 0.4) +
        geom_hline(yintercept = theta[i],linetype = 'dashed', size = 0.4)+
        theme(panel.grid = element_blank(),
              panel.background = element_blank(),
              text = element_text(size=10),
              axis.line=element_line(size = 0.3),
              axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
              axis.text = element_text(colour = "black"),
              aspect.ratio = 0.5,
              plot.margin=grid::unit(c(0,0.3,0,0), "cm")
              )
  return(pl)
}
alpha_pl <- tr_plot(pmmh2_fit$chain, 1, colours)
beta_pl <- tr_plot(pmmh2_fit$chain, 2, colours)
sigma_pl <- tr_plot(pmmh2_fit$chain, 3, colours)

pdf("Figures/Tutorial/alpha_trace.pdf")
alpha_pl
pdf("Figures/Tutorial/beta_trace.pdf")
beta_pl
pdf("Figures/Tutorial/sigma_trace.pdf")
sigma_pl
dev.off()
caption3 <- "Trace plots for each model parameter (alpha/beta/sigma resp.); generated using pmmh2()."
```

```{r, echo = F, fig.cap=caption3}
grid.arrange(
  alpha_pl,
  beta_pl,
  sigma_pl,
  nrow = 1
)
```

For $\alpha$ and $\sigma$, the traceplots in Figure 3 indicate that the chain has 
converged to its sationary regime. Another method of assessing convergence is 
through the ACF plots, which should also decay to $0$. Figure 4 shows the ACF 
plots for each parameter after taking a burn in of $2500$ iterations.

```{r include = F}
burn_in <- 2500
chain <- pmmh2_fit$chain[-(1:burn_in),]
ggplot.corr <- function(data, colour, lag.max = 100, ci = 0.95, large.sample.size = TRUE, horizontal = TRUE,...) {
  
  if(horizontal == TRUE) {numofrow <- 1} else {numofrow <- 2}
  
  list.acf <- acf(data, lag.max = lag.max, type = "correlation", plot = FALSE)
  N <- as.numeric(list.acf$n.used)
  df1 <- data.frame(lag = list.acf$lag, acf = list.acf$acf)
  df1$lag.acf <- dplyr::lag(df1$acf, default = 0)
  df1$lag.acf[2] <- 0
  df1$lag.acf.cumsum <- cumsum((df1$lag.acf)^2)
  df1$acfstd <- sqrt(1/N * (1 + 2 * df1$lag.acf.cumsum))
  df1$acfstd[1] <- 0
  df1 <- select(df1, lag, acf, acfstd)
  
  list.pacf <- acf(data, lag.max = lag.max, type = "partial", plot = FALSE)
  df2 <- data.frame(lag = list.pacf$lag,pacf = list.pacf$acf)
  df2$pacfstd <- sqrt(1/N)
    plot.acf <- ggplot(data = df1, aes( x = lag, y = acf)) +
    geom_col(color=colour, width = 0.01) +
    geom_hline(yintercept = qnorm((1+ci)/2)/sqrt(N), 
               colour = "black",
               linetype = "dashed",
               size=0.4) + 
    geom_hline(yintercept = - qnorm((1+ci)/2)/sqrt(N), 
               colour = "black",
               linetype = "dashed",
               size=0.4) + 
    scale_y_continuous(name = element_blank(), 
                       limits = c(min(df1$acf,df2$pacf),1)) +
    theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 
  plot.acf
}
acf_alpha <- ggplot.corr(chain[,1], colours[1], lag.max = 200, large.sample.size = FALSE)
acf_beta <- ggplot.corr(chain[,2], colours[2], lag.max = 200, large.sample.size = FALSE)
acf_sigma <- ggplot.corr(chain[,3], colours[3], lag.max = 200, large.sample.size = FALSE)

pdf("Figures/Tutorial/acf_alpha.pdf")
acf_alpha
pdf("Figures/Tutorial/acf_beta.pdf")
acf_beta
pdf("Figures/Tutorial/acf_sigma.pdf")
acf_sigma
dev.off()
caption4 <- "ACF plots for each model parameter (alpha/beta/sigma resp.); generated using pmmh2()."
```

```{r, echo = F, fig.cap=caption4}
grid.arrange(
  acf_alpha, 
  acf_beta,
  acf_sigma,
  nrow = 1
)
```

By lag 100, both $\alpha$ and $\sigma$'s ACF appear to have decayed sufficiently.
However, $\beta$'s ACF plot suggestes that there is still a large level of 
correlation between values by lag 200. As a result, one might consider running
the chain for longer. Finally, Figure 5 gives the histograms of the approximated
posterior distributions, which are all roughly centred on the correct values.

```{r include=F}
param = theta
samples = as.data.frame(chain)
colnames(samples) = c("alpha", "beta", "sigma")
alpha_posterior <- ggplot(data = samples)+
  geom_histogram(aes(alpha), bins = 50, color = "#264653", fill="#264653",size = 0.4,alpha=0.4)+
  geom_vline(xintercept = param[1],linetype = 'dashed',size = 0.4)+
  labs(x = expression(alpha), y="")+
  scale_x_continuous(limits=c(0,1),breaks=c(seq(0,1,by=0.2),param[1]),labels=c(seq(0,1,by=0.2),param[1]))+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 

beta_posterior <- ggplot(data = samples)+
  geom_histogram(aes(beta), bins=50, color = "#e76f51", fill="#e76f51",size = 0.4,alpha=0.4)+
  geom_vline(xintercept = param[2],linetype = 'dashed',size = 0.4)+
  labs(x = expression(beta),y="")+
  scale_x_continuous(limits=c(0.5,2.5),breaks=c(seq(0.5,2.5,by=0.25),param[3]),labels=c(seq(0.5,2.5,by=0.25),param[3]))+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        ) 

sigma_posterior <- ggplot(data = samples)+
  geom_histogram(aes(sigma), bins=50, color = "#B17D00", fill="#B17D00",size = 0.4,alpha=0.4)+
  geom_vline(xintercept = param[3],linetype = 'dashed',size = 0.4)+
  labs(x = expression(sigma),y="")+
  scale_x_continuous(limits=c(0.5,2.5),breaks=c(seq(0.5,2.5,by=0.25),param[3]),labels=c(seq(0.5,2.5,by=0.25),param[3]))+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text = element_text(size=10),
        axis.line=element_line(size = 0.3),
        axis.title.y = element_text(angle=90, vjust=0.5, hjust=0.5),
        axis.text = element_text(colour = "black"),
        aspect.ratio = 0.5,
        plot.margin=grid::unit(c(0,0.3,0,0), "cm")
        )
```

```{r include = F}
pdf("Figures/Tutorial/posterior_alpha.pdf")
alpha_posterior
pdf("Figures/Tutorial/posterior_beta.pdf")
beta_posterior
pdf("Figures/Tutorial/posterior_sigma.pdf")
sigma_posterior
caption5 <- "Sample histograms for each model parameter (alpha/beta/sigma resp.); generated using pmmh2()."

```

```{r echo=F, fig.cap=caption5}
suppressWarnings(
grid.arrange(
  alpha_posterior,
  beta_posterior,
  sigma_posterior,
  nrow = 1
)
)
```